Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2018-02-09 paper/ResponseToReviews.tex

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

\title{Response to reviews of ``Mapping mutational effects along the evolutionary landscape of HIV envelope'' for \textit{eLife}}
\author{Hugh K. Haddox, Adam S. Dingens, Sarah K. Hilton, Julie Overbaugh, and Jesse D. Bloom}

\begin{document}
\maketitle

\emph{Below, the original comments {\color{blue} are in blue}, and our responses are in black.}

\color{blue}

\section*{SUMMARY}

Information on the possible amino acid trajectories of the rapidly evolving HIV envelope protein has long been limited. In this manuscript, you extend previous work from your lab analyzing the mutational landscape of the HIV-1 Env gene during in vitro viral replication. The work focuses on two clade A transmitter founder viruses, BG505 and BF520, which differ by more than 100 mutations. A deep mutational scanning approach is used to identify the permitted mutations at each residue of Env. Of particular interest is how amino acid preferences differ for these two proteins. A set of 30 sites with clear shifts in amino acid preference are identified. These data are compared to known mutational constraints on Env and to one another, and sources for the shifts in mutant affects are proposed by eliminating many scenarios. The results offer a body of knowledge useful for identifying i) residues and networks that are critical for basic function and ii) hot spots for evolvability or immune pressure. The analysis has been performed to a high standard (though some clarifications are necessary; see minor comments), and the openness of the data is a great service to the community. The paper represents a significant contribution towards understanding the structure-function relationship of HIV envelope. Therefore, the paper is likely to be accepted for publication if the following points are adequately addressed.

{\color{black}
Thanks for the excellent summary of our work.
We appreciate the favorable evaluation of the rigor of the work and the importance of the data.
The comments below are extremely helpful, and we have revised the manuscript in line with these suggestions.
These revisions have further strengthened the manuscript.
}

\section*{ESSENTIAL REVISIONS}

1. How are mutations that strongly modulate Env's expression level taken into account?

{\color{red}
The best I can think of here is just to explain our rationale for excluding signal peptide and cytoplasmic tail. 
I don't think we have a way of addressing this comment beyond that?
Hugh and Adam, do you have additional ideas?
}

2. Only the native, pre-fusion Env trimer is used to calculate proximity of substituted and shifted residues. It is possible that substitution-shift colocalization may be detected in other conformations of Env as it undergoes structural rearrangement to mediate fusion, as opposed to acting through longer-range allostery in the pre-fusion conformation. Another well-characterized structure that should be considered is the post-fusion six-helix bundle of gp41.

{\color{red}
This is a really good suggestion.
Hugh, are you able to add analysis with this and any other relevant conformations that are substantially different?
We should think about whether these would be added just as analyses, or whether we'd also show such a structure.
}

3. Overall, it is found that the shifted sites tend to cluster together, but they are not necessarily the same as the sites with substitutions between the two Envs or ones that contact them. This analysis is not pursued in great detail, however, leaving the reasons why shifted sites tend to cluster together unexplained. Given that the authors also observe some evidence of entrenchment of substitutions between the Envs, is it possible that interactions between the shifted sites themselves are important? This question could be explored through a more detailed examination of the structure.

{\color{red}
Seems like this might be best examined by looking carefully at some of the clusters?
I'm not sure how to do it with a statistical analysis.
Hugh and Adam---do you have any thoughts on this?
}

\section*{MINOR POINTS}

1. Readers may miss the rationale for excluding the signal peptide and cytoplasmic tail if it is only found in the Methods.

{\color{red}Yes, we should move this up. Will be done in context of answering essential revision 1, and then we can describe change here.}

2. Page 3: When referring to the modifications made in this paper versus the previous publication, it is not clear whether the cell type difference was expression of CCR5 or use of a new cell line altogether.

{\color{red}
Easy to fix.
}

3. Page 6: The dN/dS parameter is introduced, but its meaning in the two different models is only described in the next paragraph. Noting that a description follows may help readers.

{\color{red}
Easy to fix.}

4. The color scheme of structures in Fig 7a, 9a is difficult to discern. Higher contrast colors please.

{\color{red}
Hugh, can you do this?}

5. The methodology used here was previously developed in your group, which would benefit from a short description in the main text. For example, how exactly is $\omega$ defined, what is the stringency parameter, can the use of $\omega$ be motivated, how is $\omega_r$ defined and motivated etc? What are the assumptions in using these measures? We think this is necessary for a stand-alone paper aimed at a broad audience. All the details of the method should not be repeated, but an intuitive summary in the main text would be helpful.

{\color{red}
This is easy to do. Maybe Jesse should do this?}

6. With the current version 2.2.0 of dms\_tools2 installed via pip today (January 19, 2018), the included iPython notebook throws an error. Error text is below: 
\begin{verbatim}
--------------------------------------------------------------------------- 
ModuleNotFoundError Traceback (most recent call last) 
in <module>() 
36 import dms_tools2.dssp 
37 import dms_tools2.compareprefs 
---> 38 import dms_tools2.protstruct 
39 import dms_tools2.rplot 
40 print(dms_tools2.rplot.versionInfo())

ModuleNotFoundError: No module named 'dms_tools2.protstruct'
\end{verbatim}

{\color{red}
This is now fixed by the latest version.}

7. It might be helpful to increase the transparency on the points in Figure 3B, because at the current level it's impossible to resolve any detail in the blob near zero.

{\color{red}
Not sure if this will work in practice because then other points become to sparse. 
Jesse can also make contour plots to show as supplement.}

8. In Figure 4, it's not clear why the p-value for $\omega$ greater/less than one is used for shading rather than the estimated $\omega$ itself. 
{\color{red}
Jesse can add explanation here.}

9. What is the rationale for rescaling amino acid preferences by the stringency parameter before comparing them? Based on (Hilton et al., 2017), my understanding is that only the amino acid preferences are measured directly. The stringency parameter appears to be based on a phylogenetic model fit. This rescaling would emphasize differences in the basic ordering of amino acid preferences over differences of intensity, but the latter would seem to also be important. 
{\color{red}
Jesse can respond to this. I suppose we could also repeat all with un-rescaled if we wanted.}

10. In the text (p8, bottom) it is stated that the amino acid preferences are rescaled for comparison. But this is not mentioned in Figure 6 or described clearly in the Methods (p18, bottom). For the sake of clarity, it would be helpful to introduce new notation for rescaled preferences and indicate clearly where they are used. 
{\color{red}
We can do this.}

11. In the discussion it is noted that: "But although there is some entrenchment of differences between BG505 and BF520, this is not the major factor behind the shifts in amino-acid preferences: most sites that have shifted between BG505 and BF520 actually have the same wildtype amino acid in both Envs." Does wildtype refer to the amino acids in the actual BG505 and BF520 sequences? The most preferred amino acid is the same in only 12/30 cases, based on Figure 6C. 
{\color{red}
Yes, it refers to wildtype. I suppose we could modify to show wildtype in plot, but not sure if that is needed.}

12. A body of computational work that has been done to infer the effects of epistasis on HIV proteins (including ENV) is not adequately referenced. (Posted 5th Feb 2018) 
{\color{red}
This includes the new Arup Chakroborty paper. Others?}

\end{document}

### 2018-02-09 paper/eLife_original_submission.tex
added response to reviews outline

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

\subsection{Deep mutational scanning of each Env}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV growth in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
Our modified deep mutational scanning strategy is in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

\begin{figure}
\begin{minipage}[t]{0.34\textwidth}
{\bf \Large A} \\ 
\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
\centerline{\small BG505}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
\centerline{\small BF520}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
\end{minipage}
\begin{minipage}[t]{0.01\textwidth}
\end{minipage}
\begin{minipage}[t]{0.65\textwidth}
{\bf \Large B} \\
\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
\end{minipage}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
}
\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in Figure~\ref{fig:mutfreqs}.
Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
{\input{figures/avgmutfreqs}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\end{figure}

The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (Figure~\ref{fig:mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in Figure~\ref{fig:mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:divselRSA}).

@@

Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:glycans}).
All sites that evolve slower than expected for both experimental datasets are in Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}.

@@

All analyses used the phylogenetic tree in Figure~\ref{fig:tree}.

@@

Relative solvent accessibilities were calculated as in Figure~\ref{fig:shiftfeatures}.

>>>>

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 1in 0in 1.4in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in \FIG{tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable were masked as indicated in \FIG{tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to other sequences at alignable sites is in \FIGSUPP[tree]{identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences to BG505 and BF520.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in \FIG{tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in \FIG{tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

@@

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
\FIG{tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity).
In our experiments, we analyze only the ectodomain and transmembrane domain of Env (we exclude the signal peptide and cytoplasmic tail).
In these regions of Env, BG505 and BF520 are identical at 549 of the 616 sites (89.1\% identity) that are alignable across clade A Envs (\FIG{tree}-source~data~\ref{figdata:alignment}, \FIG{tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

\subsection{Deep mutational scanning of each Env}
We have previously described a deep mutational scanning strategy for measuring how all amino-acid mutations to Env affect HIV growth in cell culture, and applied this strategy to the late-stage lab-adapted LAI strain~\citep{haddox2016experimental}.
Here we made several modifications to this earlier strategy to apply it to transmitted-founder Envs and to reduce the experimental noise. 
This last consideration is especially important when comparing Envs, since it is only possible to reliably detect differences that exceed the magnitude of the experimental noise. 
Our modified deep mutational scanning strategy is in \FIG{dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support growth of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (\FIG{dms_schematic}A).
We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (\FIG{dms_schematic}B).
Our libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene approximately followed a Poisson distribution with a mean between 1 and 1.5 (\FIGSUPP[dms_schematic]{fsanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV growth, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

\begin{figure}
\begin{minipage}[t]{0.34\textwidth}
{\bf \Large A} \\ 
\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
\centerline{\small BG505}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
\centerline{\small BF520}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
\end{minipage}
\begin{minipage}[t]{0.01\textwidth}
\end{minipage}
\begin{minipage}[t]{0.65\textwidth}
{\bf \Large B} \\
\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
\end{minipage}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral-replication errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (\FIGSUPP[mutfreqs]{mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in \FIG{mutfreqs}-source~data~\ref{figdata:allprefs}.
}
\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in \FIG{mutfreqs}.
Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
{\input{figures/avgmutfreqs}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\end{figure}

The deep mutational scanning effectively selected for functional Envs as evidenced by strong purifying selection against stop codons.
\FIG{mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (\FIG{mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (\FIGSUPP[mutfreqs]{mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (\FIGSUPP[mutfreqs]{mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral growth.
\FIG{mutfreqs}A only summarizes one aspect of the deep mutational scanning data, but Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in \FIG{dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in \FIG{mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, which had replicate-to-replicate Pearson correlations of only $R = 0.45$ to $0.50$.

While the replicates are well correlated across all replicates for both BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa (\FIG{mutfreqs}B, compare red and blue versus gray plots).
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.
Note also that there is a trend for highly preferred amino acids to be more strongly preferred in BG505 than BF520 (most high-preference points in the gray plots in \FIG{mutfreqs}B fall above the diagonal); however, this trend does not necessarily reflect differences between the Envs.
Rather, there were modest differences in the stringency of selection between our deep mutational scans of BG505 and BF520 (\FIGSUPP[mutfreqs]{mutfreqs} shows that purifying selection better purged stop codons in BG505).
In the next section, we correct for these experimental differences by calibrating each dataset to match the stringency of selection in nature.

@@

For both Envs, sites that evolve significantly slower or faster than expected are often on Env's surface (\FIGSUPP[divsel]{divselRSA}).

@@

Many of the sites of slower-than-expected evolution are asparagines in N-linked glycosylation motifs (\FIGSUPP[divsel]{glycans}).
All sites that evolve slower than expected for both experimental datasets are in \FIGSUPP[divsel]{slower}.

@@

All analyses used the phylogenetic tree in \FIG{tree}.

@@

Relative solvent accessibilities were calculated as in \FIG{shiftfeatures}.
